Open source symbolic multibody dynamics software

### 2012-02-13 PetersonGedeHubbard2012.tex
Fixed some typos, error in definition of B_u

\section{Derivation of linearization algorithm}
\label{derivations}
The goal of the following calculations is to concretely establish the first
order relationships (about an arbitrary equilibrium point) between an arbitary
choice of independent coordinates and speeds, and the time-derivatives of all
coordinates and speeds (both independent and dependent).  This enables several
common tasks: (1) stability analysis of arbitrary equilibria, (2) traditional
control system design, (3) control system design with output feedback based
upon quantities that are dependent (i.e., not state variables), (4)
determination of measurement equations for use in Kalman filter design when
measured quantities are at the acceleration level (i.e., accelerometers).  The
formalism presented here is generic enough to cover most examples of
time-varying constrained multibody systems with arbitrary external inputs and
arbitrary specified quantities.

@@

    $\bm{f}_{c}(\bm{q}, t)$ & $\bm{0}$ & $f_c : \mathbf{R}^n \times \mathbf{R} \to

@@

$\bm{f}_{v}(\bm{q}, \bm{u}, t)$ & $\bm{0}$ & $f_v : \mathbf{R}^n \times

@@

$f_a : \mathbf{R}^n \times \mathbf{R}^n \times \mathbf{R}^o \times

@@

Equating the right hand sides of equations (\ref{eq:f0_plus_f1_expansion}),
(\ref{eq:acceleration_expansion}),
and (\ref{eq:f2_plus_f3_expansion}) to zero (as per table
\ref{table:assumptions}), and introducing the following definitions
\begin{align}
\tilde{M}_{qq}  &\triangleq \nabla_{\bm{\dot{q}}}\bm{f}_0 \\
\tilde{M}_{uuc} &\triangleq \nabla_{\bm{\dot{u}}}\bm{f}_a \\
\tilde{M}_{uud} &\triangleq \nabla_{\bm{\dot{u}}}\bm{f}_2 \\
\tilde{A}_{qq}  &\triangleq -\nabla_{\bm{q}}(\bm{f}_0 + \bm{f}_1) \\
\tilde{A}_{qu}  &\triangleq -\nabla_{\bm{u}}\bm{f}_1 \\
\tilde{A}_{uqc} &\triangleq -\left[ \nabla_{\bm{\dot{q}}}\bm{f}_a \tilde{M}_{qq}^{-1} \tilde{A}_{qq} + \nabla_{\bm{q}} \bm{f}_a\right] \\
\tilde{A}_{uuc} &\triangleq -\left[ \nabla_{\bm{\dot{q}}}\bm{f}_a \tilde{M}_{qq}^{-1} \tilde{A}_{qu} + \nabla_{\bm{u}} \bm{f}_a\right] \\
\tilde{A}_{uqd} &\triangleq -\left[ \nabla_{\bm{\dot{q}}}\bm{f}_3 \tilde{M}_{qq}^{-1} \tilde{A}_{qq} + \nabla_{\bm{q}} (\bm{f}_2 + \bm{f}_3)\right] \\
\tilde{A}_{uud} &\triangleq -\left[ \nabla_{\bm{\dot{q}}}\bm{f}_3 \tilde{M}_{qq}^{-1} \tilde{A}_{qu} + \nabla_{\bm{u}} \bm{f}_u\right] \\
\tilde{B}_{u}   &\triangleq \nabla_{\bm{r}}\bm{f}_{3}
\end{align}
enables the unconstrained linear state space equations to be written as
\begin{align}
  \label{eq:state_space_unconstrained}
  \left[
    \begin{array}{cc}
      \tilde{M}_{qq} & \bm{0}_{n \times o} \\
      \bm{0}_{m \times n} & \tilde{M}_{uuc} \\
      \bm{0}_{(o-m) \times n} & \tilde{M}_{uud}
    \end{array}
    \right]
    \left[
      \begin{array}{c}
        \delta \bm{\dot{q}} \\
        \delta \bm{\dot{u}}
      \end{array}
    \right]
   &=
   \left[
     \begin{array}{cc}
       \tilde{A}_{qq} & \tilde{A}_{qu} \\
       \tilde{A}_{uqc} & \tilde{A}_{uuc} \\
       \tilde{A}_{uqd} & \tilde{A}_{uud}
     \end{array}
   \right]
    \left[
      \begin{array}{c}
        \delta \bm{q} \\
        \delta \bm{u}
      \end{array}
    \right]
    +
    \left[
      \begin{array}{c}
        \bm{0}_{(n + m) \times q} \\
        \tilde{B}_{u}
      \end{array}
    \right]
    \delta \bm{r}
\end{align}
Equation (\ref{eq:state_space_unconstrained}) has a state space of dimension $n
+ o$, yet only $p = n - l + o - m$ of these quantities are independent.
To address this issue, a particular set of independent
coordinates and speeds must be selected. To this end, consider the following
partitioning of the generalized coordinates and generalized speeds:
\begin{equation*}
  \tilde{\bm{q}} \triangleq \left[\begin{array}{cc}\bm{q}_{i} &
      \bm{q}_{d}\end{array}\right]^{T} =  P_{q}^{-1} \bm{q}
      \qquad\qquad
  \tilde{\bm{u}} \triangleq \left[\begin{array}{cc}\bm{u}_{i} &
      \bm{u}_{d}\end{array}\right]^{T} =  P_{u}^{-1} \bm{u}
\end{equation*}
where $P_q \in \mathbf{R}^{n \times n}$ and $P_u \in \mathbf{R}^{o \times o}$
are invertible permutation matrices which map an ordering which has the
independent quantities ($\bm{q}_{i}\in\mathbf{R}^{n-l},\,
\bm{u}_{i}\in\mathbf{R}^{o-m})$ first, followed by the dependent quantities
($\bm{q}_{d}\in\mathbf{R}^{l},\, \bm{u}_{d}\in\mathbf{R}^{m}$) to the
original ordering of the coordinates and speeds.  We use the notation $P_{qi}$
and $P_{qd}$ to denote the first $n-l$ and last $l$ columns of $P_q$,
respectively; similarly, $P_{ui}$ is the first $o-m$ colums of $P_{u}$ while
$P_{ud}$ is the last $m$ columns of $P_u$.

Setting the right hand side of equation (\ref{eq:configuration_expansion}) equal to
zero, and making use of $P_q$, we have
\begin{align}
  \nabla_{\bm{q}}\bm{f}_{h} P_{q} \cdot \delta \bm{\tilde{q}} &=
  \nabla_{\bm{q}}\bm{f}_{h} P_{qi} \cdot \delta \bm{q_i} +
  \nabla_{\bm{q}}\bm{f}_{h} P_{qd} \cdot \delta \bm{q_d}\notag\\
  &= \bm{0}\notag\\
  \implies \delta \bm{q}_d &= -(\nabla_{\bm{q}}\bm{f}_{h} P_{qd})^{-1}
  (\nabla_{\bm{q}}\bm{f}_{h} P_{qi}) \cdot \delta \bm{q}_i \notag\\
  \implies \delta \bm{q} &= \left[ I_{n \times n} - P_{qd}(\nabla_{\bm{q}}
    \bm{f}_{h} P_{qd})^{-1} \nabla_{\bm{q}} \bm{f}_{h} \right] P_{qi} \cdot \delta
    \bm{q}_i
\end{align}
For convenience, we define
\begin{equation}
  \label{eq:C_0}
  C_0 \triangleq \left[ I_{n \times n} - P_{qd}(\nabla_{\bm{q}}
    \bm{f}_{h} P_{qd})^{-1} \nabla_{\bm{q}} \bm{f}_{h} \right] P_{qi}
\end{equation}
which enables us to write
\begin{align}
  \label{eq:delta_q}
  \delta \bm{q} &= C_0 \cdot \delta \bm{q}_i
\end{align}

Applying the same approach to (\ref{eq:velocity_expansion}), we obtain
\begin{align}
\nabla_{\bm{q}}f_{nh} \cdot \delta \bm{q} +
\nabla_{\bm{u}}f_{nh} \cdot \delta \bm{u} &= \nabla_{\bm{q}} \bm{f}_{nh} \cdot
\delta \bm{q} + \nabla_{\bm{u}} \bm{f}_{nh} P_{ui} \cdot \delta \bm{u}_i +
\nabla_{\bm{u}} \bm{f}_{nh} P_{ud} \cdot \delta \bm{u}_d \notag\\
&= \bm{0}\notag\\
\implies \delta \bm{u}_d &= -\left(\nabla_{\bm{u}} \bm{f}_{nh}
P_{ud}\right)^{-1}\left[\nabla_{\bm{q}}\bm{f}_{nh} \cdot \delta\bm{q} +
  \nabla_{\bm{u}} \bm{f}_{nh} P_{ui} \cdot \delta \bm{u}_i \right]\notag\\
  \implies \delta \bm{u} &= -P_{ud}(\nabla_{\bm{u}} \bm{f}_{nh} P_{ud})^{-1}
  \nabla_{\bm{q}} \bm{f}_{nh} \cdot \delta \bm{q}\notag\\
  &+ \left[I - P_{ud} (\nabla_{\bm{u}}\bm{f}_{nh} P_{ud})^{-1} \nabla_{\bm{u}}
    \bm{f}_{nh} \right] P_{ui} \cdot \delta \bm{u}_i
\end{align}
We define
\begin{align}
  \label{eq:C_1}
  C_1 &\triangleq -P_{ud}(\nabla_{\bm{u}} \bm{f}_{nh} P_{ud})^{-1}
  \nabla_{\bm{q}} \bm{f}_{nh} \\
  \label{eq:C_2}
  C_2 &\triangleq \left[I - P_{ud} (\nabla_{\bm{u}}\bm{f}_{nh} P_{ud})^{-1} \nabla_{\bm{u}}
    \bm{f}_{nh} \right] P_{ui}
\end{align}
which enables us to write
\begin{align}
  \label{eq:delta_u}
  \delta \bm{u} &= C_1 \cdot \delta \bm{q} + C_2 \cdot \delta \bm{u}_i\notag\\
  &= C_1 C_0 \cdot \delta \bm{q}_i + C_2 \cdot \delta \bm{u}_i
\end{align}

By making use of equations (\ref{eq:delta_q}) and (\ref{eq:delta_u}), we can
rewrite equation (\ref{eq:state_space_unconstrained}) as
\begin{align}
  \label{eq:state_space_constrained}
  \left[
    \begin{array}{cc}
      \tilde{M}_{qq} & \bm{0}_{n \times o} \\
      \bm{0}_{m \times n} & \tilde{M}_{uuc} \\
      \bm{0}_{(o-m) \times n} & \tilde{M}_{uud}
    \end{array}
    \right]
    \left[
      \begin{array}{c}
        \delta \bm{\dot{q}} \\
        \delta \bm{\dot{u}}
      \end{array}
    \right]
   &=
   \left[
     \begin{array}{cc}
       (\tilde{A}_{qq} + A_{qu} C_1 ) C_0 & \tilde{A}_{qu} C_2 \\
       (\tilde{A}_{uqc} + A_{uuc} C_1 ) C_0 & \tilde{A}_{uuc} C_2\\
       (\tilde{A}_{uqd} + A_{uud} C_1 ) C_0 & \tilde{A}_{uud} C_2
     \end{array}
   \right]
    \left[
      \begin{array}{c}
        \delta \bm{q}_i \\
        \delta \bm{u}_i
      \end{array}
    \right]
    +
    \left[
      \begin{array}{c}
        \bm{0}_{(n+m) \times q} \\
        \tilde{B}_{u}
      \end{array}
    \right]
    \delta \bm{r}
\end{align}

There are two requirements for this method to work, namely, both
$\nabla_{\bm{q}}\bm{f}_{h} P_{qd}$ and $\nabla_{\bm{u}} \bm{f}_{nh} P_{ud}$
must be nonsingular.  Because of the presence of the permutation matrices, this
implies proper selection of dependent coordinates and dependent speeds is
critical.  The permutation matrices effectively select the columns of the
constraint gradient matrices, and may be selected by examination of the
constraint gradient matrix through a SVD or other numerical techniques; for
small systems it may be possible to choose dependent speeds without resorting
to numerical techniques if the system trajectories stay with the configuration
space where these matrices are nonsingular.

>>>>

\section{Derivation of linearization algorithm}
\label{derivations}
The goal of the following calculations is to concretely establish the first
order relationships (about an arbitrary equilibrium point) between an arbitrary
choice of independent coordinates and speeds, and the time-derivatives of all
coordinates and speeds (both independent and dependent).  This enables several
common tasks: (1) stability analysis, (2) control system design, (3) control system design with output feedback based
upon quantities that are dependent (i.e., not state variables), (4)
determination of measurement equations for use in Kalman filter design when
measured quantities are at the acceleration level (i.e., accelerometers).  The
formalism presented here is generic enough to cover most examples of
time-varying constrained multibody systems with arbitrary external inputs and
arbitrary specified quantities.

@@

    $\bm{f}_{c}(\bm{q}, t)$ & $\bm{0}$ & $\bm{f}_c : \mathbf{R}^n \times \mathbf{R} \to

@@

$\bm{f}_{v}(\bm{q}, \bm{u}, t)$ & $\bm{0}$ & $\bm{f}_v : \mathbf{R}^n \times

@@

$\bm{f}_a : \mathbf{R}^n \times \mathbf{R}^n \times \mathbf{R}^o \times

@@

Equating the right hand sides of equations (\ref{eq:f0_plus_f1_expansion}),
(\ref{eq:acceleration_expansion}),
and (\ref{eq:f2_plus_f3_expansion}) to zero (as per table
\ref{table:assumptions}), and introducing the following definitions
\begin{align}
\tilde{M}_{qq}  &\triangleq \nabla_{\bm{\dot{q}}}\bm{f}_0 \\
\tilde{M}_{uuc} &\triangleq \nabla_{\bm{\dot{u}}}\bm{f}_a \\
\tilde{M}_{uud} &\triangleq \nabla_{\bm{\dot{u}}}\bm{f}_2 \\
\tilde{A}_{qq}  &\triangleq -\nabla_{\bm{q}}(\bm{f}_0 + \bm{f}_1) \\
\tilde{A}_{qu}  &\triangleq -\nabla_{\bm{u}}\bm{f}_1 \\
\tilde{A}_{uqc} &\triangleq -\left[ \nabla_{\bm{\dot{q}}}\bm{f}_a \tilde{M}_{qq}^{-1} \tilde{A}_{qq} + \nabla_{\bm{q}} \bm{f}_a\right] \\
\tilde{A}_{uuc} &\triangleq -\left[ \nabla_{\bm{\dot{q}}}\bm{f}_a \tilde{M}_{qq}^{-1} \tilde{A}_{qu} + \nabla_{\bm{u}} \bm{f}_a\right] \\
\tilde{A}_{uqd} &\triangleq -\left[ \nabla_{\bm{\dot{q}}}\bm{f}_3 \tilde{M}_{qq}^{-1} \tilde{A}_{qq} + \nabla_{\bm{q}} (\bm{f}_2 + \bm{f}_3)\right] \\
\tilde{A}_{uud} &\triangleq -\left[ \nabla_{\bm{\dot{q}}}\bm{f}_3 \tilde{M}_{qq}^{-1} \tilde{A}_{qu} + \nabla_{\bm{u}} \bm{f}_3\right] \\
\tilde{B}_{u}   &\triangleq -\nabla_{\bm{r}}\bm{f}_{3}
\end{align}
enables the unconstrained linear state space equations to be written as
\begin{align}
  \label{eq:state_space_unconstrained}
  \left[
    \begin{array}{cc}
      \tilde{M}_{qq} & \bm{0}_{n \times o} \\
      \bm{0}_{m \times n} & \tilde{M}_{uuc} \\
      \bm{0}_{(o-m) \times n} & \tilde{M}_{uud}
    \end{array}
    \right]
    \left[
      \begin{array}{c}
        \delta \bm{\dot{q}} \\
        \delta \bm{\dot{u}}
      \end{array}
    \right]
   &=
   \left[
     \begin{array}{cc}
       \tilde{A}_{qq} & \tilde{A}_{qu} \\
       \tilde{A}_{uqc} & \tilde{A}_{uuc} \\
       \tilde{A}_{uqd} & \tilde{A}_{uud}
     \end{array}
   \right]
    \left[
      \begin{array}{c}
        \delta \bm{q} \\
        \delta \bm{u}
      \end{array}
    \right]
    +
    \left[
      \begin{array}{c}
        \bm{0}_{(n + m) \times q} \\
        \tilde{B}_{u}
      \end{array}
    \right]
    \delta \bm{r}
\end{align}
Equation (\ref{eq:state_space_unconstrained}) has a state space of dimension $n
+ o$, yet only $p = n - l + o - m$ of these quantities are independent.
To address this issue, a particular set of independent
coordinates and speeds must be selected. To this end, consider the following
partitioning of the generalized coordinates and generalized speeds:
\begin{equation*}
  \tilde{\bm{q}} \triangleq \left[\begin{array}{cc}\bm{q}_{i} &
      \bm{q}_{d}\end{array}\right]^{T} =  P_{q}^{-1} \bm{q}
      \qquad\qquad
  \tilde{\bm{u}} \triangleq \left[\begin{array}{cc}\bm{u}_{i} &
      \bm{u}_{d}\end{array}\right]^{T} =  P_{u}^{-1} \bm{u}
\end{equation*}
where $P_q \in \mathbf{R}^{n \times n}$ and $P_u \in \mathbf{R}^{o \times o}$
are invertible permutation matrices which map an ordering which has the
independent quantities ($\bm{q}_{i}\in\mathbf{R}^{n-l},\,
\bm{u}_{i}\in\mathbf{R}^{o-m})$ first, followed by the dependent quantities
($\bm{q}_{d}\in\mathbf{R}^{l},\, \bm{u}_{d}\in\mathbf{R}^{m}$) to the
original ordering of the coordinates and speeds.  We use the notation $P_{qi}$
and $P_{qd}$ to denote the first $n-l$ and last $l$ columns of $P_q$,
respectively; similarly, $P_{ui}$ is the first $o-m$ columns of $P_{u}$ while
$P_{ud}$ is the last $m$ columns of $P_u$.

Setting the right hand side of equation (\ref{eq:configuration_expansion}) equal to
zero, and making use of $P_q$, we have
\begin{align}
  \nabla_{\bm{q}}\bm{f}_{c} P_{q} \cdot \delta \bm{\tilde{q}} &=
  \nabla_{\bm{q}}\bm{f}_{c} P_{qi} \cdot \delta \bm{q_i} +
  \nabla_{\bm{q}}\bm{f}_{c} P_{qd} \cdot \delta \bm{q_d}\notag\\
  &= \bm{0}\notag\\
  \implies \delta \bm{q}_d &= -(\nabla_{\bm{q}}\bm{f}_{c} P_{qd})^{-1}
  (\nabla_{\bm{q}}\bm{f}_{c} P_{qi}) \cdot \delta \bm{q}_i \notag\\
  \implies \delta \bm{q} &= \left[ I_{n \times n} - P_{qd}(\nabla_{\bm{q}}
    \bm{f}_{c} P_{qd})^{-1} \nabla_{\bm{q}} \bm{f}_{c} \right] P_{qi} \cdot \delta
    \bm{q}_i
\end{align}
For convenience, we define
\begin{equation}
  \label{eq:C_0}
  C_0 \triangleq \left[ I_{n \times n} - P_{qd}(\nabla_{\bm{q}}
    \bm{f}_{c} P_{qd})^{-1} \nabla_{\bm{q}} \bm{f}_{c} \right] P_{qi}
\end{equation}
which enables us to write
\begin{align}
  \label{eq:delta_q}
  \delta \bm{q} &= C_0 \cdot \delta \bm{q}_i
\end{align}

Applying the same approach to (\ref{eq:velocity_expansion}), we obtain
\begin{align}
  \nabla_{\bm{q}}\bm{f}_{v} \cdot \delta \bm{q} +
  \nabla_{\bm{u}}\bm{f}_{v} \cdot \delta \bm{u} &= \nabla_{\bm{q}} \bm{f}_{v} \cdot
\delta \bm{q} + \nabla_{\bm{u}} \bm{f}_{v} P_{ui} \cdot \delta \bm{u}_i +
\nabla_{\bm{u}} \bm{f}_{v} P_{ud} \cdot \delta \bm{u}_d \notag\\
&= \bm{0}\notag\\
\implies \delta \bm{u}_d &= -\left(\nabla_{\bm{u}} \bm{f}_{v}
P_{ud}\right)^{-1}\left[\nabla_{\bm{q}}\bm{f}_{v} \cdot \delta\bm{q} +
  \nabla_{\bm{u}} \bm{f}_{v} P_{ui} \cdot \delta \bm{u}_i \right]\notag\\
  \implies \delta \bm{u} &= -P_{ud}(\nabla_{\bm{u}} \bm{f}_{v} P_{ud})^{-1}
  \nabla_{\bm{q}} \bm{f}_{v} \cdot \delta \bm{q}\notag\\
  &+ \left[I - P_{ud} (\nabla_{\bm{u}}\bm{f}_{v} P_{ud})^{-1} \nabla_{\bm{u}}
    \bm{f}_{v} \right] P_{ui} \cdot \delta \bm{u}_i
\end{align}
We define
\begin{align}
  \label{eq:C_1}
  C_1 &\triangleq -P_{ud}(\nabla_{\bm{u}} \bm{f}_{v} P_{ud})^{-1}
  \nabla_{\bm{q}} \bm{f}_{v} \\
  \label{eq:C_2}
  C_2 &\triangleq \left[I - P_{ud} (\nabla_{\bm{u}}\bm{f}_{v} P_{ud})^{-1} \nabla_{\bm{u}}
    \bm{f}_{v} \right] P_{ui}
\end{align}
which enables us to write
\begin{align}
  \label{eq:delta_u}
  \delta \bm{u} &= C_1 \cdot \delta \bm{q} + C_2 \cdot \delta \bm{u}_i\notag\\
  &= C_1 C_0 \cdot \delta \bm{q}_i + C_2 \cdot \delta \bm{u}_i
\end{align}

By making use of equations (\ref{eq:delta_q}) and (\ref{eq:delta_u}), we can
rewrite equation (\ref{eq:state_space_unconstrained}) as
\begin{align}
  \label{eq:state_space_constrained}
  \left[
    \begin{array}{cc}
      \tilde{M}_{qq} & \bm{0}_{n \times o} \\
      \bm{0}_{m \times n} & \tilde{M}_{uuc} \\
      \bm{0}_{(o-m) \times n} & \tilde{M}_{uud}
    \end{array}
    \right]
    \left[
      \begin{array}{c}
        \delta \bm{\dot{q}} \\
        \delta \bm{\dot{u}}
      \end{array}
    \right]
   &=
   \left[
     \begin{array}{cc}
       (\tilde{A}_{qq} + \tilde{A}_{qu} C_1 ) C_0 & \tilde{A}_{qu} C_2 \\
       (\tilde{A}_{uqc} + \tilde{A}_{uuc} C_1 ) C_0 & \tilde{A}_{uuc} C_2\\
       (\tilde{A}_{uqd} + \tilde{A}_{uud} C_1 ) C_0 & \tilde{A}_{uud} C_2
     \end{array}
   \right]
    \left[
      \begin{array}{c}
        \delta \bm{q}_i \\
        \delta \bm{u}_i
      \end{array}
    \right]
    +
    \left[
      \begin{array}{c}
        \bm{0}_{(n+m) \times q} \\
        \tilde{B}_{u}
      \end{array}
    \right]
    \delta \bm{r}
\end{align}

There are two requirements for this method to work, namely, both
$\nabla_{\bm{q}}\bm{f}_{c} P_{qd}$ and $\nabla_{\bm{u}} \bm{f}_{v} P_{ud}$
must be nonsingular.  Because of the presence of the permutation matrices, this
implies proper selection of dependent coordinates and dependent speeds is
critical.  The permutation matrices effectively select the columns of the
constraint gradient matrices, and may be selected by examination of the
constraint gradient matrix through a SVD or other numerical techniques; for
small systems it may be possible to choose dependent speeds without resorting
to numerical techniques if the system trajectories stay with the configuration
space where these matrices are nonsingular.
